---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data


We load data and reclassify date column as date format.

```{r Load, results = "hide"}
setwd("D:/Coursera/Data Science With R/Reproducible Research/Project 1")
data <- read.csv("activity.csv")
data$date <- as.Date(data$date, "%Y-%m-%d")
```


## What is mean total number of steps taken per day?


A quick summary of steps variable by date is as below:

Total number of steps taken per day

```{r sum}
with(data, tapply(steps, date, sum, na.rm = T))
```


Mean of the total number of steps taken per day

```{r mean}
meansteps <- with(data, tapply(steps, date, mean, na.rm = T))
round(meansteps, 2)
```


Median of the total number of steps taken per day

```{r median}
with(data, tapply(steps, date, median, na.rm = T))
```


## What is the average daily activity pattern?


Here, we plot 5-minute interval with average steps across all days to see overall pattern

```{r plot}
sum.byint <- with(data, tapply(steps, interval, sum, na.rm = T))
ave.byint <- sum.byint / sum(data$steps, na.rm = T)
int <- unique(data$interval)
plot(int, ave.byint, type = "l", xlab = "5-minute Interval (0 - 2355)", ylab = "Average Steps")
title(main = "Average Daily Activity Pattern")
```


```{r, echo = FALSE, results = "hide"}
maxsteps <- max(sum.byint)
row <- grep(maxsteps, sum.byint)
maxint <- names(sum.byint)[row]
```

Based on the above plot, we see that within the interval of **`r maxint`**, there is a maximum number of steps taken of **`r maxsteps`**.


## Imputing missing values


Now we deal with missing values (coded as **NA**) in the dataset.

Calculate the total number of missing values in the dataset

```{r NA}
misval <- is.na(data)
sum(misval)
```


In an attempt to fill in the missing values in the dataset, we decide to replace them with the average of steps taken across all days. We do so by the following:

```{r impute}
data$steps[is.na(data$steps)] <- mean(data$steps, na.rm = T)
newdata <- data
head(newdata)
```


We then create a histogram showing the total number of steps taken each day.

```{r histogram}
newsum <- with(newdata, tapply(steps, date, sum))
hist(newsum, breaks = 20, xlab = "Total number of steps", main = "Total number of steps taken each day")
```


The mean total number of steps taken per day of the new dataset is

```{r newmean}
newmean <- with(newdata, tapply(steps, date, mean))
round(newmean, 2)
```


The median total number of steps taken per day of the new dataset is

```{r newmedian}
newmedian <- with(newdata, tapply(steps, date, median))
round(newmedian, 2)
```

After imputing missing data, we observe that, in terms of both mean and median, only certain days with NA values have their number of steps taken changed.  


## Are there differences in activity patterns between weekdays and weekends?


Create a new factor variable with two levels - "weeday" and "weekend"

```{r newvar}
day <- with(newdata, weekdays(date))
week <- ifelse(day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
newdata <- cbind(newdata, week)
str(newdata)
```


Now we make a plot to compare daily activity pattern between weekdays and weekends

```{r compare}
int <- unique(newdata$interval)
sum.byweek <- with(newdata, tapply(steps, week, sum))
weekday <- subset(newdata, week == "Weekday", c("steps", "interval"))
weekday <- with(weekday, tapply(steps, interval, sum))
weekday <- weekday / sum.byweek[1]

weekend <- subset(newdata, week == "Weekend", c("steps", "interval"))
weekend <- with(weekend, tapply(steps, interval, sum))
weekend <- weekend / sum.byweek[2]

par(mfcol = c(2, 1), mar = c(2, 2, 2, 2))
plot(int, weekday, type = "l", col = "blue", xlab = "5-minute Interval (0 - 2355)", ylab = "Average Steps", main = "Weekdays VS Weekends")
plot(int, weekend, type = "l", col = "red", xlab = "5-minute Interval (0 - 2355)", ylab = "Average Steps")

```
